---
title: "Setup: Packages"
# yaml-author
date: 2020-04-30
output: github_document
time: 5
reading: 0
---

# Setup: Packages

*Purpose*: Every time you start an analysis, you will need to *load packages*.
This is a quick warmup to get you in this habit.

*Reading*: (None)

**Note**: If you open this `.Rmd` file in RStudio you can *execute* the chunks of code (stuff between triple-ticks) by clicking on the green arrow to the top-right of the chunk, or by placing your cursor within a chunk (between the ticks) and pressing `Ctrl + Shift + Enter`. See [here](https://rmarkdown.rstudio.com/authoring_quick_tour.html) for a brief introduction to RMarakdown files. Note that in RStudio you can Shift + Click (CMD + Click) to follow a link.

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

### __q1__ Create a new code chunk and prepare to load the `tidyverse`.

In [RStudio](https://bookdown.org/yihui/rmarkdown/r-code.html) use the shortcut
`Ctrl + Alt + I` (`Cmd + Option + I` on Mac). Type the command
`library(tidyverse`).

```{r}
library(tidyverse)
```



Make sure to load packages at the *beginning* of your notebooks! This is a
best-practice.

### __q2__ Run the chunk with `library(tidyverse)`. What packages are loaded?

✔ tibble  3.1.8      ✔ dplyr   1.0.10
✔ tidyr   1.2.1      ✔ stringr 1.4.1 
✔ readr   2.1.3      ✔ forcats 0.5.2

In
[RStudio](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts)
press `Ctrl + Alt + T` (`Cmd + Option + T` on Mac) to run the code chunk at your
cursor.


### __q3__ What are the main differences between `install.packages()` and `library()`? How often do you need to call each one?

install.packages() is used to install new packages in R, while library() is used to load packages that have already been installed.

When installing a new package, you would use install.packages("package_name"). Once the package is installed, you can load it into your R session by using library("package_name").

You would need to use install.packages() whenever you want to install a new package that you haven't used before. You would use library() whenever you want to use a package that you have previously installed and want to load into your current R session.


<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-setup01-packages-assignment.Rmd).
