---
title: "RMS Titanic"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|-------------------|-----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

In this dataset, we can see that there is class, sex, age, survived (yes or no), and n (the number of people who fit the criteria).

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

The Wikipedia page states that there were 2,224 total passengers and crew members aboard, over 1500 of whom died.

\- Are there any differences?

Yes, there is a difference of 2224-2201 = 23 people total.

\- If yes, what might account for those differences?

Some people's bodies may not have been found, so there could be a lack of information on them, or there could have been a general loss of information over time, considering how long ago the incident occurred.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>% 
  filter(Survived == "Yes") %>% 
  ggplot(aes(x = Class, y = n, fill = Sex)) +
  geom_col(position = "dodge")
```

**Observations**:

Among the passengers, more females survived (although 3rd class has almost an equal number of female and male survivors), but among the crew members, significantly more males survived.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>% 
  filter(Survived == "Yes") %>% 
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge")
```

**Observations**:

It appears as though 100% of both female and male passengers in 1st and 2nd class survived, which does not make sense with reality since we know that not all of them survived, so there is some *fishy-ness* with this aspect of the plot. With the 3rd class passengers and crew members, it appears that a greater proportion of females survived than the proportion of males.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>% 
  filter(Survived == "Yes") %>% 
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge") + 
  facet_grid(~ Age)
```

**Observations**:

This new plot shows that there were greater proportions of female survivors than proportions of male survivors among all of the adults. With the children, it appears as though 100% of all of the female and male children in 1st and 2nd class survived, a greater proportion of female children survived than male children in 3rd class, and none of the "crew" children survived, which makes sense since there are no crew members who are children! The *fishy-ness* in q4 might have been from not having accounted for the proportion of children who survived and grouping them with the general population of passengers.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
