---
title: "SAT and College Grades"
author: "Sreenidhi Chalimadugu"
date: 2023-04-19
output:
  github_document:
    toc: true
prerequisites:
  - e-vis00-basics
---

*Purpose*: How do we apply hypothesis testing to investigating data? In this challenge you'll practice using hypothesis testing tools to make sense of a dataset.

*Reading*:
- [Harvard Study Says SATs Should Be Optional: Here's Why](https://www.csmonitor.com/USA/USA-Update/2016/0120/Harvard-study-says-SATs-should-be-optional.-Here-s-why) (Optional); easy-to-read news article on colleges going SAT-free
- [Norm-Referenced Tests and Race-Blind Admissions](https://cshe.berkeley.edu/publications/norm-referenced-tests-and-race-blind-admissions-case-eliminating-sat-and-act-university) (Optional); technical report on relationship between the SAT/ACT and non-academic factors

*Credit*: This is based on a [case study](http://onlinestatbook.com/2/case_studies/sat.html) originally prepared by Emily Zitek, with data collected through the research of Thomas MacFarland.

```{r setup}
library(tidyverse)
library(readxl)
library(broom)
library(modelr)
library(rsample)
```

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

*Background*: Every year about 2 million students take the Scholastic Aptitude Test (SAT). The exam is [controversial](http://www.nea.org/home/73288.htm) but [extremely consequential](https://www.csmonitor.com/2004/0518/p13s01-legn.html). There are many claims about the SAT, but we're going to look at just one: Is the SAT predictive of scholastic performance in college? It turns out this is a fairly complicated question to assess---we'll get an introduction to some of the complexities.

# Obtain the Data
<!-- -------------------------------------------------- -->

### __q1__ Visit the [SAT and College GPA](http://onlinestatbook.com/2/case_studies/sat.html) case study page, scroll to the bottom, and click the `Open Data with Excel` button. This will allow you to download an `xls` file. Save the file to your `data` folder, load the data as `df_sat`, and perform your "first checks" against these data. Answer the questions below:

```{r q1-task}
## TODO:
df_sat <- read_excel("data/sat.xls")

## TODO: Do your "first checks"
df_sat %>% 
  glimpse()
```

**Observations**:

- Fill in the following "data dictionary"

| Column     | Meaning                             |
|------------|-------------------------------------|
| `high_GPA` | high school GPA of a student        |
| `math_SAT` | math SAT score of a student         |
| `verb_SAT` | verbal SAT score of a student       |
| `comp_GPA` | computer science GPA of a student   |
| `univ_GPA` | overall university GPA of a student |

- What information do we have about these students?
  - We have the student grades from high school, SAT scores, and grades from university.
- What kinds of information *do we not have* about these students?
  - We do not have any other performance metrics besides these GPAs and SAT scores, and we are also missing demographic information.
- Based on these missing variables, what possible effects could be present in the data that we would have *no way of detecting*?
  - Based on these missing variables, we have no way of detecting if the sample is biased or if there are other variables at play that affect the information we are presented.

# Analysis with Hypothesis Testing
<!-- ----------------------------------------------------------------------- -->

We're going to use two complementary approaches to analyze the data, the first based on hypothesis testing of correlation coefficients, and the second based on fitting a regression model and interpreting the regression coefficients.

To simplify the analysis, let's look at a composite SAT score:

```{r compute-composite}
## NOTE: No need to edit this
df_composite <-
  df_sat %>%
  mutate(both_SAT = math_SAT + verb_SAT)

df_composite
```

## View 1: Correlations
<!-- ----------------------------------------------------------------------- -->

### __q2__ Create a *single* plot that shows `univ_GPA` against *both* `high_GPA` and `both_SAT`. Visually compare the two trends.

*Hint*: One way to do this is to first *pivot* `df_composite`.

```{r q2-task}
## TODO:
df_composite %>%
  pivot_longer(
    cols = c('high_GPA', 'both_SAT'),
    names_to = 'score_type',
    values_to = 'score' ) %>%
  ggplot(aes(univ_GPA, score)) + 
  geom_point() + 
  facet_grid(rows = vars(score_type), scales = "free")
```

**Observations**:

- What relationship do `univ_GPA` and `both_SAT` exhibit?
  - There seems to be a positive correlation between univ_GPA and both_SAT, especially for university GPAs greater than 3.0.

- What relationship do `univ_GPA` and `high_GPA` exhibit?
  - Similarly, there seems to be a positive correlation between univ_GPA and high_SAT, especially for university GPAs greater than 3.0.

### Hypothesis Testing with a Correlation Coefficient
<!-- ------------------------- -->

We can use the idea of hypothesis testing with a correlation coefficient. The idea is to set our null hypothesis to the case where there is no correlation, and test to see if the data contradict that perspective. Formally, the null (H0) and alternative (HA) hypotheses relating to a correlation coefficient between
two variables `X, Y` are:

$$\text{H0: } \text{Corr}[X, Y] = 0$$

$$\text{HA: } \text{Corr}[X, Y] \neq 0$$

The R function `cor.test` implements such a hypothesis test under the assumption that `X, Y` are both normally distributed. First, let's check to see if this assumption looks reasonable for our data.

### __q3__ Plot histograms for `both_SAT, high_GPA, univ_GPA`. Which---if any---of the variables look approximately normally distributed.

```{r q3-task}
df_composite %>% 
  pivot_longer(
    cols = c('high_GPA', 'both_SAT', 'univ_GPA'),
    names_to = "metric",
    values_to = "measure"
  ) %>% 
  ggplot(aes(x = measure)) +
  geom_histogram(bins = 20) +
  facet_grid(~metric, scales = "free")
```

**Observations**:

- To what extent does `both_SAT` look like a normal distribution?
  - `both_SAT` looks more right skewed than a normal distribution.
- To what extent does `high_GPA` look like a normal distribution?
  - `high_GPA` looks more left skewed than a normal distribution, and there is also a large gap in the middle at 3.0 GPA, so it's not a very normal distribution.
- To what extent does `univ_GPA` look like a normal distribution?
  - `univ_GPA` looks very left skewed, as there are many more higher GPAs than lower ones.

Keep in mind your findings as you complete q4.

### __q4__ Use the function `cor.test()` to construct confidence intervals for `corr[high_GPA, univ_GPA` and `corr[both_SAT, univ_GPA]`. Answer the questions below.

```{r q4-task}
## TODO: Use the function cor.test() to test the correlations between
##       high_GPA and univ_GPA, as well as between
##       both_SAT and univ_GPA

cor.test(df_composite$high_GPA, df_composite$univ_GPA)
cor.test(df_composite$both_SAT, df_composite$univ_GPA)
```

**Observations**:

- Which correlations are significantly nonzero?
  - Both of the correlations are significantly nonzero (0.78 and 0.68).
- Which of `high_GPA` and `both_SAT` seems to be more strongly correlated with `univ_GPA`?
  - `high_GPA` seems to be more strongly correlated with univ_GPA, with a correlation coefficient of 0.78, as opposed to that of `both_SAT` with 0.68.
- How do the results here compare with the visual you created in q2?
  - The correlation coefficient results make sense with the visual from q2, since from the plot it appears that `univ_GPA` is more strongly correlated with `high_GPA` and it has less of a scatter overall compared to the plot with `both_SAT`.
- Based on these results, what can we say about the predictive capabilities of both `high_GPA` and `both_SAT` to predict `univ_GPA`?
  - `high_GPA` can predict `univ_GPA` slightly better than `both_SAT` can, since it is more strongly correlated given the higher correlation coefficient. 

Finally, let's use the bootstrap to perform the same test using *different* assumptions.

### __q5__ Use the bootstrap to approximate a confidence interval for `corr[high_GPA, univ_GPA`. Compare your results---both the estimate and confidence interval---to your results from q4.

```{r q5-task}
## TODO: Use the bootstrap to compute a confidence interval for corr[high_GPA, univ_GPA]
n_bootstrap <- 1000

bootstrap_cor <- numeric(n_bootstrap)
for (i in 1:n_bootstrap) {
  sample_id <- sample(length(df_composite$high_GPA), replace = TRUE)
  sample_high_GPA <- df_composite$high_GPA[sample_id]
  sample_univ_GPA <- df_composite$univ_GPA[sample_id]

  bootstrap_cor[i] <- cor(sample_high_GPA, sample_univ_GPA)
}

# Compute confidence interval
lower_conf_interval <- quantile(bootstrap_cor, 0.025)
upper_conf_interval <- quantile(bootstrap_cor, 0.975)
est <- quantile(bootstrap_cor, 0.5)

lower_conf_interval
upper_conf_interval
```

**Observations**:

- How does your estimate from q5 compare with your estimate from q4?
  - The estimate from q5 is similar to the estimate from q4.
- How does your CI from q5 compare with your CI from q4?
  - The confidence interval is the same.

*Aside*: When you use two different approximations to compute the same quantity and get similar results, that's an *encouraging sign*. Such an outcome lends a bit more credibility to the results.

## View 2: Modeling
<!-- ------------------------- -->

Correlations are useful for relating two variables at a time. To study the relationship among more variables we can instead use a fitted model. Using a model, we can also help assess whether it is *worthwhile* to measure a variable.

To begin, let's first split the data into training and validation sets.

```{r split}
## NOTE: No need to edit
set.seed(101)

df_train <-
  df_composite %>%
  rowid_to_column() %>%
  slice_sample(n = 80)

df_validate <-
  df_composite %>%
  rowid_to_column() %>%
  anti_join(
    .,
    df_train,
    by = "rowid"
  )
```

### Hypothesis Testing with a Model
<!-- ------------------------- -->

We can combine the ideas of hypothesis testing with a model. Using a model, we can express our hypotheses in terms of the model parameters. For instance, if we were interested in whether $X$ has an affect on $Y$, we might set up a model:

$$Y_i = \beta X_i + \epsilon_i$$

With the hypotheses:

$$\text{H0}: \beta = 0$$

$$\text{HA}: \beta \neq 0$$

In this case, we're testing for whether $X$ has a significant effect on $Y$. Let's apply this idea to relating the variables `univ_GPA` and `high_GPA`. Luckily R has built-in tools to construct a confidence interval on the $\beta$'s in a regression [1]; we'll simply use those tools rather than do it by hand.

### __q6__ Fit a linear model predicting `univ_GPA` with the predictor `both_SAT`. Assess the model to determine how effective a predictor `both_SAT` is for `univ_GPA`. Interpret the resulting confidence interval for the coefficient on `both_SAT`.

```{r q6-task}
## TODO: Fit a model of univ_GPA on the predictor both_SAT
fit_basic <-
  df_composite %>% 
  lm(formula = univ_GPA ~ both_SAT)

## NOTE: The following computes confidence intervals on regression coefficients
fit_basic %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )
```

**Observations**:

- What is the confidence interval on the coefficient of `both_SAT`? Is this coefficient significantly different from zero?
  - The confidence interval is around 0.0020 to 0.0035, which is significantly different from zero, since it does not actually include zero.
- By itself, how well does `both_SAT` predict `univ_GPA`?
  - It can provide a general prediction that with a higher SAT there will be a higher GPA, but this is not a very strong correlation.

Remember from `e-model03-interp-warnings` that there are challenges with interpreting regression coefficients! Let's investigate that idea further.

### __q7__ Fit a model predicting `univ_GPA` using both `high_GPA` and `both_SAT`. Compare the prediction accuracy and hypothesis test results.

```{r q7-task}
## TODO: Fit and assess models with predictors both_SAT + high_GPA, and high_GPA alone
df_composite %>% 
  lm(formula = univ_GPA ~ high_GPA + both_SAT) %>% 
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )

df_composite %>% 
  lm(formula = univ_GPA ~ high_GPA) %>% 
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )
```

**Observations**:

- How well do these models perform, compared to the one you built in q6?
  - Both models perform well.
- What is the confidence interval on the coefficient of `both_SAT` when including `high_GPA` as a predictor?? Is this coefficient significantly different from zero?
  - The confidence interval here is -0.00022 to 0.0018, which includes zero and therefore the coefficient is not significantly different from zero.
- How do the hypothesis test results compare with the results in q6?
  - Taking the `high_GPA` into account, `both_SAT` might not be positively correlated with academic performance, unlike what the results show in q6.


## Synthesize
<!-- ------------------------- -->

Before closing, let's synthesize a bit from the analyses above.

### __q8__ Using the results from all previous q's, answer the following questions.

**Observations**:

- Between `both_SAT` and `high_GPA`, which single variable would you choose to predict `univ_GPA`? Why?
  - Based on the data we have, it seems as though `high_GPA` predicts `univ_GPA` better than `both_SAT`, one reason being that is has a greater correlation coefficient with `univ_GPA` compared to `both_SAT`.
- Is `both_SAT` an effective predictor of `univ_GPA`? What specific pieces of evidence do you have in favor of `both_SAT` being effective? What specific pieces of evidence do you have against?
  - According to the data, it does not seem like `both_SAT` is an effective predictor of `univ_GPA`. Given that there initially seems to be a positive correlation between `both_SAT` and `univ_GPA`, it might make it seem like it is an effective predictor of `univ_GPA`, but taking `high_GPA` into account, `both_SAT` does not positively correlate with `univ_GPA`.

# End Notes
<!-- ----------------------------------------------------------------------- -->

[1] There are also assumptions underlying this kind of testing, for more information see this [Wiki article](https://en.wikipedia.org/wiki/Linear_regression#Assumptions).
